\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2014-10-14 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
All edits complete.

This goal is formalized using model scoring from Bayesian statistics. In its raw form this
almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix} we
derived the simplified Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
If multiple splits do, whichever maximizes the right side is chosen.

@@

To model properly, hBOA is designed to work with large population sizes, resulting in a large number of
evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
terminated after performing generations equal to $N$.

@@

\subsection{Local Search}
Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
best of all techniques on Deceptive Trap, Deceptive Step Trap, HIFF, Ising Spin Glass, and MAX-SAT problems.
$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
this method is not enough to quickly reach the global optima in many of these problems which causes it to eventually
be overtaken by the model building techniques.

@@

Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
integration makes it improve much sooner. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
P3 is able to immediately integrate optimal versions of each
trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.

\section{Computational Expenses}
While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
complexity, in some domains this will not be true. Model based methods are especially likely to violate
this norm. Therefore in order to assess P3's quality in solving problems with efficient fitness functions,
we provide data on both its algorithmic complexity and wall clock time.

@@

To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
of 100 runs, with unsuccessful runs treated as slower than successful runs.
In total these runs
took over 6 computing years to complete. We were therefore forced to use
diverse hardware potentially running multiple processes in parallel.
As such these timings should be considered very skeptically as a number
of sources may contribute to biased error in the results. However, due to
the size of the dataset and the general regularity in the results general conclusions
can still be made.

\subsubsection{Model Building}
When considering wall clock time hBOA and Parameter-less hBOA perform significantly worse than when using
evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while under reasonable
assumptions P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
As P3 and LTGA require a similar asymptotic complexity per
evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.

\subsubsection{P3}
Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
On
Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT it still finds the global
optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
than LTGA but by no more than a constant factor.

@@

To achieve high success rates on randomly generated problem classes LTGA and hBOA must use a population size
large enough to solve the hardest instances in that class. Therefore these methods will have
population sizes larger than necessary to solve the easiest instances in the class. Even
on single instance problems, both methods will require population sizes large enough to
ensure the worst random initialization is diverse enough to solve the problem, which may be
much larger than the best random initialization.

@@

When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
Counterintuitively the use of hill climbing on the initial population reduces
P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
makes its aggregate crossover success incomparable to P3's.

Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
crossover P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
and quickly fell behind P3 in intermediate fitness quality. Therefore even the infrequently successful crossover donation are
critical to success. This does however suggest a potential avenue for future improvement by using more successful
modeling and donation algorithms.

@@

Another feature unique to P3 worth investigating is the shape and size of the population pyramid
constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
each level of the pyramid for the largest tested problem size for each problem. Each point is the
median size across 100 runs. If a run did not store any solutions at a level it is treated as 0,
and no point is drawn if the median run had 0 solutions stored at that level. While pyramid size
is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
is representative of that for all other tested problem sizes.

With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
in size as the level increases. This is because a solution must be a strict fitness improvement
over its previously added version to be added to a higher level, which becomes less likely
each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
population size decreases each generation of traditional evolutionary search.
By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
search on high quality solutions at higher levels. In comparison,
LTGA and hBOA suboptimally use a fixed population size at each generation.

>>>>

This goal is formalized using model scoring from Bayesian statistics. In its raw form this
almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
If multiple splits do, whichever maximizes the right side is chosen.

@@

hBOA is designed to work with large population sizes, resulting in a large number of
evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
terminated after performing generations equal to $N$.

@@

\subsection{Local Search}
Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
this method is slow in reaching the global optima in many of these problems which causes it to eventually
be overtaken by the model building techniques.

@@

Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
P3 is able to immediately integrate optimal versions of each
trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.

\section{Computational Expenses}
While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
complexity, in some domains this will not be true. Model based methods are especially likely to violate
this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
we provide data on both its algorithmic complexity and wall clock time.

@@

To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
of 100 runs, with unsuccessful runs treated as slower than successful runs.
In total these runs
took over 6 computing years to complete. We were therefore forced to use
diverse hardware potentially running multiple processes in parallel.
As such these timings should be considered very skeptically as a number
of sources may contribute to biased error in the results. However, due to
the size of the dataset and the general regularity in the results some general conclusions
can still be made.

\subsubsection{Model Building}
hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
As P3 and LTGA require a similar asymptotic complexity per
evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.

\subsubsection{P3}
Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
On
Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
than LTGA but by no more than a constant factor.

@@

To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
large enough to solve the hardest instances in that class. Therefore these methods will have
population sizes larger than necessary to solve the easiest instances in the class. Even
on single instance problems, both methods will require population sizes large enough to
ensure the worst random initialization is diverse enough to solve the problem, which may be
much larger than the best random initialization.

@@

When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
Counterintuitively, the use of hill climbing on the initial population reduces
P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
makes its aggregate crossover success incomparable to P3's.

Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
modeling and donation algorithms.

@@

Another feature unique to P3 is the shape and size of the population pyramid
constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
each level of the pyramid for the largest tested problem sizes. Each point is the
median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
is representative of that for all other tested problem sizes.

With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
in size as the level increases. This is because a solution must be a strict fitness improvement
over its previous version to be added to a higher level, which becomes less likely
each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
population size decreases each generation of traditional evolutionary search.
By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
search on high quality solutions at higher levels. In comparison,
LTGA and hBOA suboptimally use a fixed population size at each generation.
